Scheduling Musical Events in Max/MSP with Scheme for Max

### 2021-06-24 paper-1/paper-1.tex
saving work

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Scheduling Musical Events as Procedures with Scheme for Max}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Iain C.T. Duncan}
%%\authornote{Both authors contributed equally to this research.}
\email{iainctduncan@gmail.com}
\affiliation{%
  \institution{University of Victoria}
  \streetaddress{}
  \city{Victoria}
  \state{BC}
  \country{Canada}
  \postcode{}
}

@@

%% TODO visual example of a Max patch with various inputs %%

@@

Internally, Max messages are data entities consisting of a symbol that acts
as a Smalltalk-style message selector and an optional array of Max atoms. 
Each atom entity contains a member for the atom type and another for its value, 
where the type may be any of \textbf{int,} \textbf{float,} or \textbf{symbol}. 
Note that while the message selector symbol is always present at the C level,
in the visual patcher the selector may be implicit and hidden from the user. 
(i.e., the message originating from a message box that appears to contain only an integer
will actually consist of the selector ``int" followed by an atom storing the numerical value.)
The symbol ``bang" is a special symbol that can be used for a one-element message 
that essentially means ``run", and the act of triggering execution by sending a bang
message from the \textbf{bang} object is called ``banging" in the nomenclature.
Taken together, this means that in Max there are five kinds of messages: 
\textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
(The bang message is technically a symbol message
of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
%% cite puckette CMJ winter 2002 %%

@@

As an example, in the patcher screen-shot below we see a \textbf{message-box} object that will update
its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
\textbf{set hello world}. Near it is a circular \textbf{bang} object, which when clicked will send the
\textbf{message-box} the \textbf{bang} message, causing it to output the message \textbf{list hello world}.
This message will be received by the \textbf{zl.len} object, which counts the number of elements 
in any lists it receives, immediately outputting the count. The three columns in the screen-shot
show the patch as it is prior to any clicks, as it is after clicking \textbf{message-box} 1,
and as it is after clicking both \textbf{message-box} 1 and the \textbf{bang}.

%% figure one, a sample Max patch

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime, possible because s7 is implemented entirely
in ANSI C and is statically linked during compilation of S4M. 
This means the \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler". An example of this could be to control
the Max transport mechanism at a certain music time, based on data in a table 
that is used as a shared data store between multiple patcher objects.

@@

To evaluate Scheme code dynamically from the Max patcher, two input facilities
are available. If an s4m object receives a Max message consisting of the 
message selector \textbf{eval-string} and a single symbol atom, 
this is taken as a request to evaluate the symbol argument as Scheme code. 
Thus in Max, the user can build messages
with Scheme syntax or receive code as strings over the network, and by
using Max objects to convert this into one single quoted symbol, prepended
with the selector symbol \textbf{eval-string}, the code can be evaluated dynamically.
This enables users to add run-time code that is possibly long 
(i.e., too long to conveniently type into a Max \textbf{message-box}) by
sending it from a text editor or command line utility to Max over the 
local network.

The second facility for dynamic code evaluation in S4M is more unusual.
In a Max external, there exists a binding option to catch any previously
uncaught messages, so that a message with an unrecognized selector 
can be handled by a generic dispatching method. In S4M, this is used to 
catch any message that is not already reserved, with any additional 
arguments passed in as a Max list of atoms.  The \textbf{s4m} object interprets
the message to mean \textit{``evaluate this selector plus its additional atoms 
as if it is a Scheme list enclosed by parentheses"}.
Because both Max and Scheme uses white-space as separators and very little
in the way of diacritical syntax, this enables a wide variety
of simple Scheme calls to be made with little ceremony. The
individual atoms of the Max message are converted to Scheme tokens,
assembled into a list, and this list is evaluated. While seemingly simple,
this facility is of significant utility to the Max programmer, as combined
with the Max message interpolation facility, it enables the user to very
easily assemble Max widgets that trigger Scheme functions. Most anything that
can be expressed in a single s-expression without inner nesting can be used.
In the figure below,
we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
variable interpolation (the \textbf{\$1}). 
Clicking the \textbf{bang} above the \textbf{number-box} will result in a message of 
\textbf{set! foobar 74} being sent to the \textbf{s4m} object, which will then call
into Scheme with the code \textbf{(set! foobar 74)}.

%% TODO Max patch of making number message

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
 As this
thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
events. \
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Scheduling Musical Events in Max/MSP with Scheme for Max}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Iain C.T. Duncan}
%%\authornote{Both authors contributed equally to this research.}
\email{iainctduncan@gmail.com}
\affiliation{
  \institution{University of Victoria}
  \streetaddress{}
  \city{Victoria}
  \state{BC}
  \country{Canada}
  \postcode{}
}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.5\linewidth]{fig-1}
  \caption{A Max patch with a keyboard value transposed by a dial and sent to MIDI output.}
  \Description{}
\end{figure}

@@

Internally, Max messages are data entities consisting of a symbol that acts
as a Smalltalk-style message selector and an optional array of Max atoms. 
Each atom entity contains a member for the atom type and another for its value, 
where the type may be any of \textbf{int,} \textbf{float,} or \textbf{symbol}. 
Note that while the message selector symbol is always present at the C level,
in the visual patcher the selector may be implicit and hidden from the user. 
(i.e., the message originating from a message box that appears to contain only an integer
will actually consist of the selector ``int" followed by an atom storing the numerical value.)
The symbol ``bang" is a special symbol that can be used for a one-element message 
that essentially means ``run", and the act of triggering execution by sending a bang
message from the \textbf{bang} object is called ``banging" in the nomenclature.
Taken together, this means that in Max there are five kinds of messages: 
\textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
(The bang message is technically a symbol message
of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
\cite{Puckette2002}

@@

As an example, in the patcher screen-shot above we see a \textbf{message-box} object that will update
its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
\textbf{set hello world}. Near it is a circular \textbf{bang} object, which when clicked will send the
\textbf{message-box} the \textbf{bang} message, causing it to output the message \textbf{list hello world}.
This message will be received by the \textbf{zl.len} object, which counts the number of elements 
in any lists it receives, immediately outputting the count. The three columns in the screen-shot
show the patch as it is prior to any clicks, as it is after clicking \textbf{message-box} 1,
and as it is after clicking both \textbf{message-box} 1 and the \textbf{bang}.

\begin{figure}[H]
  \centering
  \includegraphics[width=.5\linewidth]{fig-2}
  \caption{Message flow at various stages}
  \Description{.}
\end{figure}

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime, possible because s7 is implemented entirely
in ANSI C and is statically linked during compilation of S4M. 
This means the \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler". In the example below, a Scheme function
is scheduled that will read a rehearsal mark from the table \textbf{mark-table},
moving the global transport to that position in 4 bars time. As the table
is a shared data object, other Max objects (or patchers) could write to this
rehearsal mark table without awareness of S4M.

\begin{figure}[H]
  \centering
  \includegraphics[width=.75\linewidth]{fig-3-transport}
  \caption{S4M patch that moves the transport in 4 bars time to a mark read from a table}
  \Description{.}
\end{figure}

@@

To evaluate Scheme code dynamically from the Max patcher, two input facilities
are available. If an s4m object receives a Max message consisting of the 
message selector \textbf{eval-string} and a single symbol atom, 
this is taken as a request to evaluate the symbol argument as Scheme code. 
Thus in Max, the user can build messages
with Scheme syntax or receive code as strings over the network, and by
using Max objects to convert this into one single quoted symbol, prepended
with the selector symbol \textbf{eval-string}, the code can be evaluated dynamically.
This enables users to add run-time code that is possibly long 
(i.e., too long to conveniently type into a Max \textbf{message-box}) by
sending it from a text editor or command line utility to Max over the 
local network. In figure 4, the left hand side shows code being sent
from a \textbf{text} object, prepared as an \textbf{eval-string} message, and sent to
the \textbf{s4m} object. A \textbf{udpreceive} object also receives code as strings
on network port 7777. (The author uses a Python script triggered
from a macro to send blocks of Scheme code from the Vim editor as
over port 7777.)

The second facility for dynamic code evaluation in S4M is more unusual.
In a Max external, there exists a binding option to catch any previously
uncaught messages, so that a message with an unrecognized selector 
can be handled by a generic dispatching method. In S4M, this is used to 
catch any message that is not already reserved, with any additional 
arguments passed in as a Max list of atoms.  The \textbf{s4m} object interprets
the message to mean \textit{``evaluate this selector plus its additional atoms 
as if it is a Scheme list enclosed by parentheses"}.
Because both Max and Scheme uses white-space as separators and very little
in the way of diacritical syntax, this enables a wide variety
of simple Scheme calls to be made with little ceremony. The
individual atoms of the Max message are converted to Scheme tokens,
assembled into a list, and this list is evaluated. While seemingly simple,
this facility is of significant utility to the Max programmer, as combined
with the Max message interpolation facility, it enables the user to very
easily assemble Max widgets that trigger Scheme functions. Most anything that
can be expressed in a single s-expression without inner nesting can be used.
In figure 4, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
variable interpolation (the \textbf{\$1}). 
Clicking the \textbf{bang} above the \textbf{number-box} or changing the number in 
the box will result in a message of 
\textbf{set-volume X} being sent to the \textbf{s4m} object, which will then call
into Scheme with the code \textbf{(set-volume X)}.

\begin{figure}[H]
  \centering
  \includegraphics[width=.75\linewidth]{fig-4-code-sends}
  \caption{S4M patch with 3 ways of sending Scheme code to the interpreter}
  \Description{.}
\end{figure}

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
events. 
% TODO: maybe a citation here??
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

% TODO figure of future events
